Incremental λ-Calculus in Cache-Transfer Style

### 2020-01-21 popl18/cts-intro.tex
Add lhs2TeX output for archival

\makeatother
\EndFmtInput
%
%
%
% First, let's redefine the forall, and the dot.
%
%
% This is made in such a way that after a forall, the next
% dot will be printed as a period, otherwise the formatting
% of `comp_` is used. By redefining `comp_`, as suitable
% composition operator can be chosen. Similarly, period_
% is used for the period.
%
\ReadOnlyOnce{forall.fmt}%
\makeatletter

@@

\part{Incremental λ-Calculus}
\label{part:incr}

\chapter{Introduction to differentiation}
\label{sec:intro}
\label{ch:static-diff-intro}

Incremental computation (or incrementalization) has a long-standing history in
computer science~\citep{Ramalingam93}.
Often, a program needs to update quickly the output of some nontrivial function $f$
when the input to the computation changes. In this scenario, we assume we have
computed \ensuremath{\Varid{y}_{1}\mathrel{=}\Varid{f}\;\Varid{x}_{1}} and we need to compute \ensuremath{\Varid{y}_{2}} that equals \ensuremath{\Varid{f}\;\Varid{x}_{2}}.
In this scenario, programmers typically have to choose between a few undesirable
options.
\begin{itemize}
\item Programmers can call again function \ensuremath{\Varid{f}} on the updated input \ensuremath{\Varid{x}_{2}} and
  repeat the computation from scratch. This choice guarantees
  correct results and is easy to implement, but typically wastes
  computation time. Often, if the updated input is close to the
  original input, the same result can be computed much faster.
\item Programmers can write by hand a new function \ensuremath{\Varid{df}} that updates the
  output based on input changes, using various techniques.
  Running a hand-written function \ensuremath{\Varid{df}} can be much more efficient than rerunning
  \ensuremath{\Varid{f}}, but writing \ensuremath{\Varid{df}} requires significant developer effort, is
  error-prone, and requires updating \ensuremath{\Varid{df}} by hand to keep it consistent with \ensuremath{\Varid{f}}
  whenever \ensuremath{\Varid{f}} is modified. In practice, this complicates code maintenance
  significantly~\citep{Salvaneschi13reactive}.
\item Programmers can write \ensuremath{\Varid{f}} using domain-specific languages that
  support incrementalization, for tasks where such languages are
  available. For instance, build scripts (our \ensuremath{\Varid{f}}) are written in
  domain-specific languages that support (coarse-grained)
  incremental builds. Database query languages also have often
  support for incrementalization.\pg{Mention here limits?}
\item Programmers can attempt using general-purpose techniques for
  incrementalizing programs, such as \emph{self-adjusting
    computation} and variants such as \emph{Adapton}. Self-adjusting
  computation applies to arbitrary purely functional programs and
  has been extended to imperative programs; however, it only
  guarantees efficient incrementalization when applied to base
  programs that are \emph{designed} for efficient
  incrementalization.\pg{Citations}
  Nevertheless, self-adjusting computation enabled incrementalizing programs
  that had never been incrementalized by hand before.
\end{itemize}

\pg{Resume and readd this text.}
% To understand how to compute |f| incrementally, we can summarize the key idea
% behing many incrementalization approaches.\pg{self-adjusting computation.}\pg{?}
% Let us assume, for simplicity, our function |f| is written in a purely
% functional language. During a computation such as |y = f x1|, each computation
% step produce an output using some inputs. The new output can in turn be used as
% input by further steps. We can record these computation steps as a directed
% acyclic graph (DAG) representing dependencies: each node is either an initial
% input or the output of some computation steps, and each output node has incoming
% edges from all

\pg{Continue discussing dependencies minimization and the
  relation with parallelism. Build scripts might be a good
  example.}

No approach guarantees automatic efficient incrementalization for arbitrary
programs.
We propose instead to design domain-specific languages
(DSLs) that can be efficiently incrementalized, that we call \emph{incremental}
DSLs (IDSLs).

To incrementalize IDSL programs, we use a transformation that we call
\emph{(finite) differentiation}.
Differentiation produces programs in the same language, called derivatives,
that can be optimized further and compiled to efficient code.
Derivatives represent changes to values through further values, that we call
simply changes.

For primitives, IDSL designers must specify the result of
differentiation: IDSL designers are to choose primitives that encapsulate
efficiently incrementalizable computation schemes, while IDSL users are to
express their computation using the primitives provided by the IDSL\@.

Helping IDSL designers to incrementalize primitives automatically is a
desirable goal, though one that we leave open. In our setting, incrementalizing
primitives becomes a problem of \emph{program synthesis}, and we agree with
\citet{Shah2017synthesis} that it should be treated as such. Among others,
\citet{Liu00} develops a systematic approach to this synthesis problem for
first-order programs based on equational reasoning, but it is unclear how
scalable this approach is. We provide foundations for using equational
reasoning, and sketch an IDSL for handling different sorts of collections. We
also discuss avenues at providing language plugins for more fundamental
primitives, such as algebraic datatypes with structural recursion.

\pg{rewrite}
In the IDSLs we consider, similarly to database languages, we use primitives for
high-level operations, of complexity similar to SQL operators.
On the one hand, IDSL designers wish to design few general primitives to limit
the effort needed for manual incrementalization.
On the other hand, overly general primitives can be harder to incrementalize
efficiently. Nevertheless, we also provide some support for more general
building blocks such as product or sum types and even (in some situations)
recursive types.
Other approaches provide more support for incrementalizing primitives, but even
then ensuring efficient incrementalization is not necessarily easy. Dynamic
approaches to incrementalization are most powerful: they can find work that can
be reused at runtime using memoization, as long as the computation is structured
so that memoization matches will occur. Moreover, for some programs it seems
more efficient to detect that some output can be reused thanks to a description
of the input changes, rather than through runtime detection.
\pg{Consider list insertion and map. This point might need to be moved to later.}
%There is a tension between the efficiency
%We return to this point in \cref{sec:fw-primitives}.
\pg{Nevertheless, we do not compete with SAC and Adapton?}

We propose that IDSLs be higher-order, so that primitives can be parameterized
over functions and hence highly flexible, and purely functional, to enable more
powerful optimizations both before and after differentiation.
Hence, an incremental DSL is a higher-order purely functional language, composed
of a $\lambda$-calculus core extended with base types and primitives.
Various database query languages support forms of finite differentiation (see
\cref{sec:finite-diff}), but only over first-order languages, which provide only
restricted forms of operators such as \ensuremath{\Varid{map}}, \ensuremath{\Varid{filter}} or aggregation.

To support higher-order IDSLs, we define the first form of differentiation that
supports higher-order functional languages; to this end, we introduce the
concept of function changes, which contain changes to either the code
of a function value or the values it closes over. While higher-order programs can be
transformed to first-order ones, incrementalizing resulting programs is still
beyond reach for previous approaches to differentiation (see
\cref{sec:finite-diff} for earlier work and \cref{sec:rw-partial-differentials}
for later approaches).
In \cref{ch:cts,ch:defunc-fun-changes} we transform higher-order programs to first-order ones by
closure conversion or defunctionalization, but we incrementalize defunctionalized programs using
similar ideas, including changes to (defunctionalized) functions.

% Instead of extending differentiation to higher-order programs, it might be
% possible to transform higher-order programs to first-order ones and try to apply
% differentiation to the result, but as
% However, efficient handling of the generated
% algebraic data types, including the ones generated by remains currently
% However, it does not appear that the resulting
% first-order programs are particularly simpler to handle.

% While there are multiple ways to transform higher-order programs to first-order
% programs where functions are represented as data, it is useful
% \pg{why not
%   defunctionalize/closure convert/...?}

\pg{actually argue for higher-order collection DSLs over relational databases;
  we didn't do that in the cited thesis section.}
%
Our primary example will be DSLs for operations on collections:
as discussed earlier (\cref{sec:aosd13-intro}), we favor
higher-order collection DSLs over relational databases.
% moving data to relational databases requires transforming them to
% a rather different metamodel
%
\pg{ What I'm saying... should also hopefully apply to data
  layout transformations used in databases.
Flattening? Nested data? Sharding?
  And data layout transformations *are*
  available in LMS... but not in GPLs?}
% We will discuss later why we favor this
% approach.\pg{where?}

We build our incremental DSLs based on
simply-typed $\lambda$-calculus (STLC), extended with
\emph{language plugins} to define the domain-specific parts, as discussed in
\cref{sec:intro-stlc} and summarized in \cref{fig:lambda-calc}. We call our
approach \emph{ILC} for \emph{Incremental Lambda Calculus}.

The rest of this chapter is organized as follows.
In \cref{sec:generalize-fin-diff} we explain that
differentiation generalizes the calculus of finite differences, a relative of
differential calculus.
In \cref{sec:motiv-example} we show a motivating example for
our approach.
In \cref{sec:change-intro} we introduce informally
the concept of \emph{changes} as values, and in \cref{sec:higher-order-intro} we
introduce \emph{changes to functions}.
In \cref{sec:informal-derive} we define differentiation and motivate it
informally.
In \cref{sec:derive-example} we apply differentiation to our motivating example.

Correctness of ILC is far from obvious. In
\cref{ch:derive-formally,ch:change-theory}, we introduce a formal theory of
changes, and we use it to formalize differentiation and prove it correct.
\pg{check later this TOC}

\section{Generalizing the calculus of finite differences}
\label{sec:generalize-fin-diff}
% Revise terminology.
Our theory of changes generalizes an existing field of mathematics called
the \emph{calculus of finite difference}: If \ensuremath{\Varid{f}} is a real
function, one can define its \emph{finite difference}, that is a
function \ensuremath{\Delta f} such that \ensuremath{\Delta f\;\Varid{a}\;\Varid{da}\mathrel{=}\Varid{f}\;(\Varid{a}\mathbin{+}\Varid{da})\mathbin{-}\Varid{f}\;\Varid{a}}. Readers
might notice the similarity (and the differences) between the
finite difference and the derivative of \ensuremath{\Varid{f}}, since the latter is
defined as
\[f'(a) = \lim_{\Varid{da} \to 0} \frac{f (a + \Varid{da}) - f(a)}{\Varid{da}}.\]

The calculus of finite differences helps computing a closed
formula for \ensuremath{\Delta f} given a closed formula for \ensuremath{\Varid{f}}. For instance,
if function \ensuremath{\Varid{f}} is defined by \ensuremath{\Varid{f}\;\Varid{x}\mathrel{=}\mathrm{2}\cdot\Varid{x}}, one can prove
its finite difference is \ensuremath{\Delta f\;\Varid{x}\;\Varid{dx}\mathrel{=}\mathrm{2}\cdot(\Varid{x}\mathbin{+}\Varid{dx})\mathbin{-}\mathrm{2}\cdot\Varid{x}\mathrel{=}\mathrm{2}\cdot\Varid{dx}}.

Finite differences are helpful for incrementalization because
they allow computing functions on updated inputs based on results
on base inputs, if we know how inputs change. Take again for
instance \ensuremath{\Varid{f}\;\Varid{x}\mathrel{=}\mathrm{2}\cdot\Varid{x}}: if \ensuremath{\Varid{x}} is a base input and \ensuremath{\Varid{x}\mathbin{+}\Varid{dx}}
is an updated input, we can compute \ensuremath{\Varid{f}\;(\Varid{x}\mathbin{+}\Varid{dx})\mathrel{=}\Varid{f}\;\Varid{x}\mathbin{+}\Delta f\;\Varid{x}\;\Varid{dx}}. If we already computed \ensuremath{\Varid{y}\mathrel{=}\Varid{f}\;\Varid{x}} and reuse the result, we
can compute \ensuremath{\Varid{f}\;(\Varid{x}\mathbin{+}\Varid{dx})\mathrel{=}\Varid{y}\mathbin{+}\Delta f\;\Varid{x}}. Here, the input change is
\ensuremath{\Varid{dx}} and the output change is \ensuremath{\Delta f\;\Varid{x}\;\Varid{dx}}.

However, the calculus of finite differences is usually defined
for real functions. Since it is based on operators \ensuremath{\mathbin{+}} and \ensuremath{\mathbin{-}},
it can be directly extended to commutative groups.
Incrementalization based on finite differences for groups and
first-order programs has already been
researched~\citep{Paige82FDC,GlucheGrust97Incr}, most recently and
spectacularly with DBToaster~\citep{Koch10IQE,Koch2016incremental}.

But it is not immediate how to generalize finite differencing
beyond groups. And many useful types do not form a group: for
instance, lists of integers don't form a group but only a monoid.
Moreover, it's hard to represent list changes simply through a
list: how do we specify which elements were inserted (and where),
which were removed and which were subjected to change themselves?

In ILC, we generalize the calculus of finite differences by
using distinct types for base values and changes, and adapting
the surrounding theory. ILC generalizes operators \ensuremath{\mathbin{+}} and \ensuremath{\mathbin{-}} as operators
\ensuremath{\oplus } (pronounced ``oplus'' or ``update'') and \ensuremath{\ominus } (pronounced
``ominus'' or ``difference''). We show how ILC subsumes groups in
\cref{sec:change-structure-groups}.

\section{A motivating example}
\label{sec:motiv-example}
In this section, we illustrate informally incrementalization on a
small example.

In the following program, \ensuremath{\Varid{grandTotal}\;\Varid{xs}\;\Varid{ys}} sums integer numbers in
input collections \ensuremath{\Varid{xs}} and \ensuremath{\Varid{ys}}.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{22}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\Varid{grandTotal}{}\<[22]%
\>[22]{}\mathrel{:\mkern-1mu:}\Conid{Bag}\;\mathbb{Z}\to \Conid{Bag}\;\mathbb{Z}\to \mathbb{Z}{}\<[E]%
\\
\>[3]{}\Varid{s}{}\<[22]%
\>[22]{}\mathrel{:\mkern-1mu:}\mathbb{Z}{}\<[E]%
\\[\blanklineskip]%
\>[3]{}\Varid{grandTotal}\;\Varid{xs}\;\Varid{ys}{}\<[22]%
\>[22]{}\mathrel{=}\Varid{sum}\;(\Varid{merge}\;\Varid{xs}\;\Varid{ys}){}\<[E]%
\\
\>[3]{}\Varid{s}{}\<[22]%
\>[22]{}\mathrel{=}\Varid{grandTotal}\;\Varid{xs}\;\Varid{ys}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

This program computes output \ensuremath{\Varid{s}} from input collections \ensuremath{\Varid{xs}} and
\ensuremath{\Varid{ys}}. These collections are multisets or \emph{bags}, that is,
collections that are unordered (like sets) where elements are
allowed to appear more than once (unlike sets). In this example,
we assume a language plugin that supports a base type of integers
\ensuremath{\mathbb{Z}} and a family of base types of bags \ensuremath{\Conid{Bag}\;\tau} for any type
\ensuremath{\tau}.

We can run this program on specific inputs \ensuremath{\Varid{xs}_{1}\mathrel{=}\{\mskip1.5mu \{\mskip1.5mu \mathrm{1},\mathrm{2},\mathrm{3}\mskip1.5mu\}\mskip1.5mu\}}
and \ensuremath{\Varid{ys}_{1}\mathrel{=}\{\mskip1.5mu \{\mskip1.5mu \mathrm{4}\mskip1.5mu\}\mskip1.5mu\}} to obtain output \ensuremath{\Varid{s}_{1}}. Here, double braces
\ensuremath{\{\mskip1.5mu \{\mskip1.5mu \ldots\mskip1.5mu\}\mskip1.5mu\}} denote a bag containing the elements among the double
braces.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{22}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\Varid{s}_{1}{}\<[22]%
\>[22]{}\mathrel{=}\Varid{grandTotal}\;\Varid{xs}_{1}\;\Varid{ys}_{1}{}\<[E]%
\\
\>[22]{}\mathrel{=}\Varid{sum}\;\{\mskip1.5mu \{\mskip1.5mu \mathrm{1},\mathrm{2},\mathrm{3},\mathrm{4}\mskip1.5mu\}\mskip1.5mu\}\mathrel{=}\mathrm{10}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

This example uses small inputs for simplicity, but in practice they
are typically much bigger; we use \ensuremath{\Varid{n}}
to denote the input size. In this case the asymptotic complexity of computing
\ensuremath{\Varid{s}} is \ensuremath{\Theta(\Varid{n})}.

Consider computing updated output \ensuremath{\Varid{s}_{2}} from updated inputs \ensuremath{\Varid{xs}_{2}\mathrel{=}\{\mskip1.5mu \{\mskip1.5mu \mathrm{1},\mathrm{1},\mathrm{2},\mathrm{3}\mskip1.5mu\}\mskip1.5mu\}}
and \ensuremath{\Varid{ys}_{2}\mathrel{=}\{\mskip1.5mu \{\mskip1.5mu \mathrm{4},\mathrm{5}\mskip1.5mu\}\mskip1.5mu\}}. We could recompute \ensuremath{\Varid{s}_{2}} from scratch as
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\Varid{s}_{2}{}\<[16]%
\>[16]{}\mathrel{=}\Varid{grandTotal}\;\Varid{xs}_{2}\;\Varid{ys}_{2}{}\<[E]%
\\
\>[16]{}\mathrel{=}\Varid{sum}\;\{\mskip1.5mu \{\mskip1.5mu \mathrm{1},\mathrm{1},\mathrm{2},\mathrm{3},\mathrm{4},\mathrm{5}\mskip1.5mu\}\mskip1.5mu\}\mathrel{=}\mathrm{16}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
But if the size of the updated inputs is \ensuremath{\Theta(\Varid{n})}, recomputation also
takes time \ensuremath{\Theta(\Varid{n})}, and we would like to obtain our result asymptotically faster.

To compute the updated output \ensuremath{\Varid{s}_{2}} faster, we assume the changes to the
inputs have a description of size \ensuremath{\Varid{dn}} that is asymptotically smaller than the
input size \ensuremath{\Varid{n}}, that is \ensuremath{\Varid{dn}\mathrel{=}o(\Varid{n})}. All approaches to incrementalization
require small input changes. Incremental computation will then process the input
changes, rather than just the new inputs.

\section{Introducing changes}
\label{sec:change-intro}
To talk about how the differences between old values and new
values, we introduce a few concepts, for now without full definitions.
In our approach to
incrementalization, we describe changes to values as values
themselves: We call such descriptions simply \emph{changes}.
Incremental programs examine changes to inputs to understand how to produce
changes to outputs. Just
like in STLC we have terms (programs) that evaluates to values,
we also have \emph{change terms}, which evaluate to \emph{change
  values}. We require that going from old values to new values
preserves types: That is, if an old value \ensuremath{\Varid{v}_{1}} has type \ensuremath{\tau},
then also its corresponding new value \ensuremath{\Varid{v}_{2}} must have type \ensuremath{\tau}.
To each type \ensuremath{\tau} we associate a type of changes or \emph{change type}
\ensuremath{\Delta \tau}: a change between \ensuremath{\Varid{v}_{1}} and \ensuremath{\Varid{v}_{2}} must be a value of type \ensuremath{\Delta \tau}.
Similarly, environments can change: to typing context \ensuremath{\Gamma} we associate
change typing contexts \ensuremath{\Delta \Gamma}, such that we can have an environment change
\ensuremath{\D\rho\typcolon\Eval{\Delta \Gamma}} from \ensuremath{\rho_{1}\typcolon\Eval{\Gamma}} to \ensuremath{\rho_{2}\typcolon\Eval{\Gamma}}.

Not all descriptions of changes are meaningful,
so we also talk about \emph{valid} changes. Valid changes satisfy additional
invariants that are useful during incrementalization.
%
A change value \ensuremath{\Varid{dv}} can be a valid change from \ensuremath{\Varid{v}_{1}} to \ensuremath{\Varid{v}_{2}}. We
can also consider a valid change as an edge from \ensuremath{\Varid{v}_{1}} to \ensuremath{\Varid{v}_{2}} in
a graph associated to \ensuremath{\tau} (where the vertexes are values of
type \ensuremath{\tau}), and we call \ensuremath{\Varid{v}_{1}} the source of \ensuremath{\Varid{dv}} and \ensuremath{\Varid{v}_{2}} the
destination of \ensuremath{\Varid{dv}}. We only talk of source and destination for valid changes:
so a change from \ensuremath{\Varid{v}_{1}} to \ensuremath{\Varid{v}_{2}} is (implicitly) valid.
We'll discuss examples of valid and invalid
changes in \cref{ex:valid-bag-int,ex:invalid-nat}.\pg{What about
  changes with multiple valid sources?}

We also introduce an operator \ensuremath{\oplus } on values and changes: if
\ensuremath{\Varid{dv}} is a valid change from \ensuremath{\Varid{v}_{1}} to \ensuremath{\Varid{v}_{2}}, then \ensuremath{\Varid{v}_{1}\oplus \Varid{dv}}
(read as ``\ensuremath{\Varid{v}_{1}} updated by \ensuremath{\Varid{dv}}'') is guaranteed to return \ensuremath{\Varid{v}_{2}}.
If \ensuremath{\Varid{dv}} is \emph{not} a valid change from \ensuremath{\Varid{v}_{1}}, then \ensuremath{\Varid{v}_{1}\oplus \Varid{dv}} can be
defined to some arbitrary value or not, without any effect on correctness. In
practice, if \ensuremath{\oplus } detects an invalid input change it can trigger an error or
return a dummy value; in our formalization we assume for simplicity that
\ensuremath{\oplus } is total.
Again, if \ensuremath{\Varid{dv}} is not valid from \ensuremath{\Varid{v}_{1}} to \ensuremath{\Varid{v}_{1}\oplus \Varid{dv}}, then we do not talk of
the source and destination of \ensuremath{\Varid{dv}}.

We also introduce operator \ensuremath{\ominus }: given two values \ensuremath{\Varid{v}_{1},\Varid{v}_{2}}
for the same type, \ensuremath{\Varid{v}_{2}\ominus \Varid{v}_{1}} is a valid change from \ensuremath{\Varid{v}_{1}}
to \ensuremath{\Varid{v}_{2}}.

Finally, we introduce change composition: if \ensuremath{\Varid{dv}_{1}} is a valid
change from \ensuremath{\Varid{v}_{1}} to \ensuremath{\Varid{v}_{2}} and \ensuremath{\Varid{dv}_{2}} is a valid change from \ensuremath{\Varid{v}_{2}} to
\ensuremath{\Varid{v}_{3}}, then \ensuremath{\Varid{dv}_{1}\circledcirc\Varid{dv}_{2}} is a valid change from \ensuremath{\Varid{v}_{1}} to
\ensuremath{\Varid{v}_{3}}.

Change operators are overloaded over different types.
Coherent definitions of validity and of operators \ensuremath{\oplus ,\ominus } and
\ensuremath{\circledcirc } for a type \ensuremath{\tau} form a \emph{change structure} over values of type
\ensuremath{\tau} (\cref{def:change-structure}).
For each type \ensuremath{\tau} we'll define a change structure (\cref{def:chs-types}),
and operators will have types \ensuremath{\oplus \typcolon\tau\to \Delta \tau\to \tau}, \ensuremath{\ominus \typcolon\tau\to \tau\to \Delta \tau}, \ensuremath{\circledcirc \typcolon\Delta \tau\to \Delta \tau\to \Delta \tau}.

\begin{example}[Changes on integers and bags]
  \label{ex:valid-bag-int}
  \label{ex:chs-int}
 \pg{changes on bags?}
To show how incrementalization affects our example, we next
describe valid changes for integers and bags.
%
For now, a change
\ensuremath{\Varid{das}} to a bag \ensuremath{\Varid{as}_{1}} simply contains all elements to be added to
the initial bag \ensuremath{\Varid{as}_{1}} to obtain the updated bag \ensuremath{\Varid{as}_{2}} (we'll ignore
removing elements for this section and discuss it later). In our
example, the change from \ensuremath{\Varid{xs}_{1}} (that is \ensuremath{\{\mskip1.5mu \{\mskip1.5mu \mathrm{1},\mathrm{2},\mathrm{3}\mskip1.5mu\}\mskip1.5mu\}}) to \ensuremath{\Varid{xs}_{2}}
(that is \ensuremath{\{\mskip1.5mu \{\mskip1.5mu \mathrm{1},\mathrm{1},\mathrm{2},\mathrm{3}\mskip1.5mu\}\mskip1.5mu\}}) is \ensuremath{\Varid{dxs}\mathrel{=}\{\mskip1.5mu \{\mskip1.5mu \mathrm{1}\mskip1.5mu\}\mskip1.5mu\}}, while the change
from \ensuremath{\Varid{ys}_{1}} (that is \ensuremath{\{\mskip1.5mu \{\mskip1.5mu \mathrm{4}\mskip1.5mu\}\mskip1.5mu\}}) to \ensuremath{\Varid{ys}_{2}} (that is \ensuremath{\{\mskip1.5mu \{\mskip1.5mu \mathrm{4},\mathrm{5}\mskip1.5mu\}\mskip1.5mu\}}) is
\ensuremath{\Varid{dys}\mathrel{=}\{\mskip1.5mu \{\mskip1.5mu \mathrm{5}\mskip1.5mu\}\mskip1.5mu\}}.
%
To represent the output change \ensuremath{\Varid{ds}} from
\ensuremath{\Varid{s}_{1}} to \ensuremath{\Varid{s}_{2}} we need integer changes. For now, we
represent integer changes as integers, and define \ensuremath{\oplus } on
integers as addition: \ensuremath{\Varid{v}_{1}\oplus \Varid{dv}\mathrel{=}\Varid{v}_{1}\mathbin{+}\Varid{dv}}.
\end{example}

For both bags and integers, a change \ensuremath{\Varid{dv}} is always valid between
\ensuremath{\Varid{v}_{1}} and \ensuremath{\Varid{v}_{2}\mathrel{=}\Varid{v}_{1}\oplus \Varid{dv}}; for other changes, however,
validity will be more restrictive.
\begin{example}[Changes on naturals]
  \label{ex:invalid-nat}
For instance, say we want to
define changes on a type of natural numbers, and we still want to
have \ensuremath{\Varid{v}_{1}\oplus \Varid{dv}\mathrel{=}\Varid{v}_{1}\mathbin{+}\Varid{dv}}. A change from \ensuremath{\mathrm{3}} to \ensuremath{\mathrm{2}} should
still be \ensuremath{\mathbin{-}\mathrm{1}}, so the type of changes must be \ensuremath{\mathbb{Z}}. But the
result of \ensuremath{\oplus } should still be a natural, that is an integer
\ensuremath{\geq \mathrm{0}}: to ensure that \ensuremath{\Varid{v}_{1}\oplus \Varid{dv}\geq \mathrm{0}} we need to require
that \ensuremath{\Varid{dv}\geq \mathbin{-}\Varid{v}_{1}}. We use this requirement to define validity on
naturals: \ensuremath{\validfromto{\mathbb{N}}{\Varid{v}_{1}}{\Varid{dv}}{\Varid{v}_{1}\mathbin{+}\Varid{dv}}} is defined as equivalent to
\ensuremath{\Varid{dv}\geq \mathbin{-}\Varid{v}_{1}}. We can guarantee equation \ensuremath{\Varid{v}_{1}\oplus \Varid{dv}\mathrel{=}\Varid{v}_{1}\mathbin{+}\Varid{dv}}
not for all changes, but only for valid changes. Conversely, if a
change \ensuremath{\Varid{dv}} is invalid for \ensuremath{\Varid{v}_{1}}, then \ensuremath{\Varid{v}_{1}\mathbin{+}\Varid{dv}\mathbin{<}\mathrm{0}}. We then
define \ensuremath{\Varid{v}_{1}\oplus \Varid{dv}} to be \ensuremath{\mathrm{0}}, though any other definition on
invalid changes would work.\footnote{In fact, we could leave
  \ensuremath{\oplus } undefined on invalid changes. Our original
  presentation~\citep{CaiEtAl2014ILC}, in essence, restricted
  \ensuremath{\oplus } to valid changes through dependent types, by ensuring
  that applying it to invalid changes would be ill-typed. Later,
  \citet{Huesca2015incrementality}, in similar developments,
  simply made \ensuremath{\oplus } partial on its domain instead of
  restricting the domain, achieving similar results.}
\end{example}
\pg{bags with removals? where?}

\subsection{Incrementalizing with changes}
After introducing changes and related notions, we describe how we incrementalize
our example program.

>>>>

\makeatother
\EndFmtInput
%
% Emacs, this is -*- latex -*-!

%
%
% First, let's redefine the forall, and the dot.
%
%
% This is made in such a way that after a forall, the next
% dot will be printed as a period, otherwise the formatting
% of `comp_` is used. By redefining `comp_`, as suitable
% composition operator can be chosen. Similarly, period_
% is used for the period.
%
\ReadOnlyOnce{forall.fmt}%
\makeatletter

@@

\section{Introduction}

Incremental computation is often desirable: after computing an output from some
input, we often need to produce new outputs corresponding to changed inputs. One
can simply rerun the same \emph{base program} on the new input; but instead,
incremental computation transforms the input change to an output change. This
can be desirable because more efficient.

Incremental computation could also be desirable because the changes themselves
are of interest: Imagine a typechecker explaining how some change to input
source propagates to changes to the typechecking result. More generally, imagine
a program explaining how some change to its input propagates through a
computation into changes to the output.

ILC (Incremental $\lambda$-Calculus)~\citep{CaiEtAl2014ILC} is
a recent approach to incremental computation for higher-order languages.
ILC represents changes from an old value \ensuremath{\Varid{v}_{1}} to an updated value \ensuremath{\Varid{v}_{2}} as a
first-class value written \ensuremath{\Varid{dv}}. ILC also transforms statically \emph{base programs}
to \emph{incremental programs} or \emph{derivatives}: derivatives produce
output changes from changes to all their inputs. Since functions are first-class
values, ILC introduces a notion of function changes.

However, as mentioned by \citeauthor{CaiEtAl2014ILC} and as we explain below,
ILC as currently defined does not allow reusing intermediate results created by the
base computation during the incremental computation. That restricts ILC to
supporting efficiently only \emph{self-maintainable computations}, a rather
restrictive class: for instance, mapping self-maintainable functions on a sequence is
self-maintainable, but dividing numbers isn't! In this paper, we remove this
limitation.

% non-blinded
% In this paper, we address this restriction as we promised in previous work.
%In keeping with our approach, we try to avoid forcing unnecessary overhead onto
%the computation.

To remember intermediate results, many incrementalization approaches rely on
forms of memoization: one uses hashtables to memoize function results, or
dynamic dependence graphs~\citep{Acar05} to remember the computation trace.
However, such data structures often remember results that might not be reused;
moreover, the data structures themselves (as opposed to their values) occupy
additional memory, looking up intermediate results has a cost in time, and
typical general-purpose optimizers cannot predict results from memoization
lookups. Instead, ILC aims to produce purely functional programs that are
suitable for further optimizations.

We eschew memoization: instead, we transform programs to
\emph{cache-transfer style (CTS)}, following ideas from \citet{Liu95}. CTS functions
output \emph{caches} of intermediate results along with their primary results. Caches
are just nested tuples whose structure is derived from code, and accessing them
does not involve looking up keys depending on inputs. We also extend
differentiation to produce \emph{CTS derivatives}, which can extract from caches
any intermediate results they need.
%\pg{they can also extract inputs, but I can't explain it here.}
This approach was inspired and pioneered by \citeauthor{Liu95} for untyped
first-order functional languages, but we integrate
it with ILC and extend it to higher-order typed languages.

% \pg{we don't!}
%and then use program transformation techniques to eliminate unneeded state.
%
While CTS functions still produce additional intermediate data structures,
produced programs can be subject to further optimizations.
We believe static analysis of a CTS function and its CTS derivative can identify
and remove unneeded state (similar to what has been done by \citeauthor{Liu95}),
as we discuss in \cref{sec:cache-pruning}.
We leave a more careful analysis to future work.

We prove most of our formalization correct in Coq
To support non-simply-typed programs, all our proofs
are for untyped $\lambda$-calculus, while previous ILC correctness proofs were
restricted to simply-typed $\lambda$-calculus.
Unlike previous ILC proofs, we simply define which changes are valid
via a \emph{logical relation}, then show the fundamental property for this
logical relation (see \cref{sec:ilc-background}). To extend this proof to
untyped $\lambda$-calculus, we switch to \emph{step-indexed} logical relations.

To support differentiation on our case studies, we also represent function changes
as closures that can be inspected, to support manipulating them more efficiently
and detecting at runtime when a function change is \emph{nil} hence need not be
applied. To show this representation is correct, we also use closures in our
mechanized proof.

Unlike plain ILC, typing programs in CTS is challenging, because the shape of
caches for a function depends on the function implementation.
Our case studies show how to non-trivially embed resulting programs in typed
languages, at least for our case studies, but our proofs support an untyped
target language.

In sum, we present the following contributions:
\pg{revise more, add sections; this is getting redundant with the text.}
\begin{itemize}
\item via examples, we motivate extending ILC to remember intermediate
  results (\cref{sec:cts-motivation});
\item we give a novel proof of correctness for ILC for untyped
  $\lambda$-calculus, based on step-indexed logical relations
  (\cref{sec:sound-derive});
\item building on top of ILC-style differentiation, we show how to transform
  untyped higher-order programs to \emph{cache-transfer-style}~(CTS)~%
  (\cref{sec:transformation});
\item we show that programs and derivatives in cache-transfer style
  \emph{simulate} correctly their non-CTS variants (\cref{sec:transformation-soundness});
\item we mechanize in Coq most of our proofs%
\begin{poplForPopl}
% PG: I'm commenting this out, assuming that we fix this soon
%, save for a few straightforward lemmas
 (see supplementary material)%
\end{poplForPopl}
;
\item we perform performance case studies (in \cref{sec:case-studies}) applying
  (by hand) extension of this technique to Haskell programs, and incrementalize
  efficiently also programs that do not admit self-maintainable derivatives.
\end{itemize}

\pg{Try merging with contributions if we can.}
The rest of the paper is organized as follows. \Cref{sec:cts-motivation} summarizes
ILC and motivates the extension to cache-transfer style.
\Cref{sec:formalization} presents our formalization and proofs.
\iftoggle{poplForThesis}{
\Cref{sec:case-studies} presents our case studies and benchmarks.
\Cref{sec:cts-limitations} discusses limitations and future work.
}{
\Cref{sec:case-studies} discusses our case studies, benchmarks,
limitations and future work.
}
\Cref{sec:cts-rw}
discusses related work and \cref{sec:conclusions} concludes.
